---
title: "Variable EDA - Jing Kai"
author: "Jing Kai Ong"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

```{r}
students <- read_csv("../data/student-mat.csv")
summary(students)
```

It does not look like there is missing data for our variables

## Part 1: Exploratory Data Analysis

### Dependent variables

We first look at the distribution of the alcohol consumption variables, `Dalc` and `Walc`

We also have the table information

```{r}

table(students$Dalc)
table(students$Walc)

ggplot(data=students,mapping=aes(x=Dalc)) +
  geom_bar(alpha = 0.75) +
  geom_bar(aes(x=Walc),alpha=0.5) +
  theme_bw()
```

It looks like there is quite a difference between the workday vs weekend alcohol consumption.

### Independent Variables

The following code provides a summary of the variables, as well as a bar plot comparing the distributions of different categories and how many lie in each category of alcohol consumption.


higher - Describes whether the student intends to pursue higher education or not (yes or no)
This variable could be significant, because we could predict that students who intend to pursue higher education are less likely to drink

```{r}
table(students$higher)

ggplot(data=students, mapping=aes(x=Dalc)) +
  geom_bar(aes(fill=higher),position="dodge2")

ggplot(data=students, mapping=aes(x=Walc)) +
  geom_bar(aes(fill=higher),position="dodge2")
```

internet - Describes whether the student has internet access at home (yes or no)

```{r}
table(students$internet)

ggplot(data=students, mapping=aes(x=Dalc)) +
  geom_bar(aes(fill=internet),position="dodge2")

ggplot(data=students, mapping=aes(x=Walc)) +
  geom_bar(aes(fill=internet),position="dodge2")

```

romantic - Describes whether the student is in a romantic relationship or not (yes or no)

```{r}
table(students$romantic)

ggplot(data=students, mapping=aes(x=Dalc)) +
  geom_bar(aes(fill=romantic),position="dodge2")

ggplot(data=students, mapping=aes(x=Walc)) +
  geom_bar(aes(fill=romantic),position="dodge2")
```

famrel - Describes the quality of the student's family relationship (scale of 1 to 5)

```{r}
table(students$famrel)

ggplot(data=students, mapping=aes(x=Dalc)) +
  geom_bar(aes(fill=as.factor(famrel)),position="dodge2")

ggplot(data=students, mapping=aes(x=Walc)) +
  geom_bar(aes(fill=as.factor(famrel)),position="dodge2")
```

freetime - Describes how much free time the student has (scale of 1 to 5)
```{r}
table(students$freetime)

ggplot(data=students, mapping=aes(x=Dalc)) +
  geom_bar(aes(fill=as.factor(freetime)),position="dodge2")

ggplot(data=students, mapping=aes(x=Walc)) +
  geom_bar(aes(fill=as.factor(freetime)),position="dodge2")

```


